Fast Sorting on Modern Computers

### 2005-07-28 docs/tech_report/shell.tex
finished shell

Shell begins by considering the array to be made up of $N/2$ subarrays of
length two, where the elements of the subarray are spaced $N/2$ keys apart.
These subarrays are then sorted using insertion sort. The first iteration will
be a simple exchange, but the keys which are exchanged move across the array by
$N/2$ spaces. The sort then considers $N/4$ arrays, who's elements are spaced
$N/4$ keys apart, which are individually sorted by insertion sort, then $N/8$
and so on. The final iteration will be a standard insertion sort, after which
the array will be sorted. A problem with this implementation is that since the
increments are powers of two, keys with odd indices are not compared with keys
with even indices until the final insertion sort. This can lengthen the final
insertion sort's running time to $O(N^2)$.

Since first published, there has been much discussion about the best
\textit{increments} to use in shellsort. Shell's original sort, described above,
has the increments $N/2$, $N/4$, $N/8$, $N/16$ ........ $1$. \cite{Sedgewick96}
discusses many increments used, and provides code for comparing these. Based on
running the \cc{driver} program\footnote{See
http://www.cs.princeton.edu/\textasciitilde{}rs/shell/}, Gonnet's shellsort was
seen to be the fastest on the authors local machine. As a result, this was used
to test the performance of shellsort. This uses increments which are
five-elevenths of the previous increment.

@@

@@

Shellsort's simulation results have the characterist shape and range of an
$O(NlogN)$ sort, albeit, not a very good one. It performs better than a
heapsort, but considerably worse than quicksort. However, it easier to program
than either of these two sort.

Shellsort performs roughly double the number of branches and instructions of
quicksort, and has more than twice as many branch mispredictions and nearly
four times as many cache misses. It's cycle count is over three times that of
quicksort, though it is more than 50 percent faster than the best of the
heapsorts.a

%TODO
TODO are there log5_11 N misses
Regenerate data before looking at this

\section{Predictor results}

>>>>

Shell begins by considering the array to be made up of $\frac{N}{2}$ subarrays
of length two, where the elements of the subarray are spaced $\frac{N}{2}$ keys
apart.  These subarrays are then sorted using insertion sort. The first
iteration will be a simple exchange, but the keys which are exchanged move
across the array by $\frac{N}{2}$ spaces. The sort then considers $\frac{N}{4}$
arrays, whose elements are spaced $\frac{N}{4}$ keys apart, which are
individually sorted by insertion sort, then $\frac{N}{8}$ and so on. The final
iteration will be a standard insertion sort, after which the array will be
sorted. A problem with this implementation is that since the increments are
powers of two, keys with odd indices are not compared with keys with even
indices until the final insertion sort. This can lengthen the final insertion
sort's running time to $O(N^2)$.

Since first published, there has been much discussion about the best
\textit{increments} to use in shellsort. Shell's original sort, described above,
has the increments $\frac{N}{2}$, $\frac{N}{4}$, $\frac{N}{8}$, $\frac{N}{16}$
........ $1$. \cite{Sedgewick96} discusses many increments used, and provides
code for comparing these. Based on running the \cc{driver} program\footnote{See
http://www.cs.princeton.edu/\textasciitilde{}rs/shell/}, Gonnet's shellsort was
seen to be the fastest on the authors local machine. As a result, this was used
to test the performance of shellsort. This uses increments which are
$\frac{5}{11}$ths of the previous increment.

@@

Improved shellsort should have a slighlty lower instruction count, and hence a
lower cycle count, than shellsort. It's cache performance should be the same,
and it should have the same number of comparitive branches, and hence a similar
number of mispredictions, though its branch count should be higher than
shellsort.

@@

Shellsort's simulation results have the characterist shape and range of an
$O(NlogN)$ sort, albeit, not a very good one. The slopes of its graphs are
slighlty steeper, as can be seen in Figures \ref{shell cycles}, \ref{shell
instruction count}, \ref{shell level 1 misses} and \ref{shell branch
mispredictions}. It performs better than a
heapsort, but considerably worse than quicksort. However, it easier to program
than either of these two sort.

Shellsort performs roughly double the number of branches and instructions of
quicksort, and has more than twice as many branch mispredictions and nearly
four times as many cache misses. It's cycle count is over three times that of
quicksort, though it is more than 50 percent faster than the best of the
heapsorts.

Shellsort's cache misses are certainly in the range of an $O(NlogN)$ sort, and
are not considerably worse than cache heapsort, and are in fact better than base
heapsort. This occurs without any cache improvements being performed upon them.
Shellsort's branch mispredictions, are, as predicted $O(Nlog_{\frac{5}{11}}N)$.

Improved shellsort does improve on shellsort, having approximately 10\% fewer
instructions. It also has slightly fewer cache misses and branch mispredictions.
Overall, this leads to a reduction in cycle count of between 35 and 70\%, as
seen in Figure \ref{shell cycle count}.

%TODO
\section{Predictor results}
